---
title: "01_Munging" 
author: "Jessica Stephens"
date: "10/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Importing Data

data is in a microsoft excel workbook
readxl

```{r load, echo=FALSE}

#check wd and file path match
setwd("C:/Users/jesse/OneDrive/Documents/Work Material/USAID/Treatment")

#read xlsx - specify path, sheet name to read

df <- read_xlsx("Data/HIV2020Estimates_ART by single year_8Sept2020_norow.xlsx",
                sheet="DataList") 


```


## Check Data Import


```{r load, echo=FALSE}

glimpse(df)
View(df)

```




## Clean dataset: Isolate Age from E_Ind and drop unnecessary columns/variables

Goal: the numbers for each one-year age band of people with HIV and of people on ART
# make a new variable with the the values after "Age" and before ";"
```{r load, echo=FALSE}



df_full_seperate <- df %>% 
    separate(E_Ind, c("A","B", "C", "D", "E"), sep = "([+;])") %>%#"B" holds the age
    separate(B, c("X", "Y", "Age_num"), sep = "([ ])")%>%
    separate(A, c("X1", "PLVHIV_or_ART", "X2"), sep = "([-])")%>%
    separate("PLVHIV_or_ART", c("PLVHIV_or_ART", "Y1"), sep = "([0])") %>% #drop 0 from PLHIV vs ART
    rename(Country=E_Count, sex=D, sex2=E)  %>%#rename country variable and vars with sex
    mutate(malefemale=ifelse(is.na(sex2), 0, 1)) %>%
    mutate(sex=ifelse(malefemale==1, "MaleFemale", sex)) %>% # not working to change sex here %>%
    mutate(Country=recode(Country, "Cote dIvoire"="Cote d'Ivoire", 
                          "Lao People Democratic Republic"="Laos",
                          "United Republic of Tanzania"= "Tanzania",
                          "Viet Nam"="Vietnam"))%>%
    select(Country, Age_num, sex, Time, Value) %>% #drop X1, X2, Y, X, C,Y1, sex2 Rounded
    view() %>%
    glimpse()
  

df_full_seperate$Age_num <-
    as.numeric(as.character(unlist(df_full_seperate$Age_num)))

view(df_full_seperate)
glimpse(df_full_seperate)

```



#############################################
# Restructure data
##############################################

## add variables for Age Categories for ACTUAL Age

Create categories for PEPFAR age brackets
<1
1-4
5-9
10-14
15-19
20-24
25-29
30-34
35-39
40-44
45-49
50+


https://stackoverflow.com/questions/12979456/r-code-to-categorize-age-into-group-bins-breaks 


```{r load, echo=FALSE}
#find max
range(df_full_seperate$Age_num)
#0-80

library(data.table)

PEPFARagebreaks19 <- c(0,1,5,10,15,20,25,30,35,40,45,50,85)
PEPFARagelabels19 <- c("<01","01--04", "05--09","10--14","15-19","20-24","25-29","30-34",
               "35-39","40-44","45-49","50+")


df_full_PEPFARage2019 <- setDT(df_full_seperate) [ , ACTUAL_Age_cat_SemiFine := cut(Age_num, 
                                breaks = PEPFARagebreaks19, 
                                right = FALSE, 
                                labels = PEPFARagelabels19)]


view(df_full_PEPFARage2019)

glimpse(df_full_PEPFARage2019)

names(df_full_PEPFARage2019)

```




# Figure out predicted ages 
#ex, 2017 ages if 2016 aged w/o other loss/gains


```{r load, echo=FALSE}

df_full_PREDICTagetime <- df_full_PEPFARage2019 %>%
 mutate(Age_predict=Age_num+1, Time_predict=Time+1)

view(df_full_PREDICTagetime)



df_full_PREDICTagegroups <- setDT(df_full_PREDICTagetime) [ , PREDICT_Age_cat_SemiFine := cut(Age_predict, 
                                breaks = PEPFARagebreaks19, 
                                right = FALSE, 
                                labels = PEPFARagelabels19)]


view(df_full_PREDICTagegroups)
glimpse(df_full_PREDICTagegroups)
#Rows: 119,280
#Columns: 11
names(df_full_PREDICTagegroups)




```



<!-- ############################################################################## -->
<!-- Multiple possible data formats, see archive for 1 & 2 -->
<!-- 1) long with disaggregate (<15 no disaggregate) -->
<!-- 2) long without sex disaggregate -->
<!-- 3) long with all data (male, female and malefemale) - can be disaggregated later -->




###########################################################
######    ALL DATA (format 3) ######
###########################################################


Multiple data formats
1) long with disaggregate
2) long without sex disaggregate
3) long with all data (male, female and malefemale) - can be disaggregated later


## Group by Age Groups and Country and sex
#actual and predicted separately
#merge actual and predicted

```{r load, echo=FALSE}

#insert the df_long_sexdisagg

#actual
df_long_Grouped_Actual_SemiFine_disagg<- df_full_PREDICTagegroups %>%
  group_by(Country, sex, Time, ACTUAL_Age_cat_SemiFine) %>%
  summarise(Total_Value_Actual=sum(Value))

glimpse(df_long_Grouped_Actual_SemiFine_disagg)
view(df_long_Grouped_Actual_SemiFine_disagg)

#predict
df_long_Grouped_Predict_SemiFine_disagg<- df_full_PREDICTagegroups %>%
  group_by(Country, sex, Time_predict, PREDICT_Age_cat_SemiFine) %>%
  summarise(Total_Value_Predict=sum(Value))

glimpse(df_long_Grouped_Predict_SemiFine_disagg)
view(df_long_Grouped_Predict_SemiFine_disagg)


#rename predict variables to match actual so they can merge
df_long_Grouped_Predict_SemiFine_vars_disagg<- df_long_Grouped_Predict_SemiFine_disagg %>%
  rename(Time=Time_predict, ACTUAL_Age_cat_SemiFine=PREDICT_Age_cat_SemiFine)

glimpse(df_long_Grouped_Predict_SemiFine_vars_disagg)
view(df_long_Grouped_Predict_SemiFine_vars_disagg)

#merge back together
df_long_Grouped_SemiFine_disagg <- df_long_Grouped_Actual_SemiFine_disagg %>% 
  left_join(df_long_Grouped_Predict_SemiFine_vars_disagg, by=c("Time"="Time", "ACTUAL_Age_cat_SemiFine"="ACTUAL_Age_cat_SemiFine","Country"="Country", "sex"="sex"))

view(df_long_Grouped_SemiFine_disagg)
# 
# write_csv(df_long_Grouped_SemiFine_disagg,"C:/Users/jesse/OneDrive/Documents/Work Material/USAID/Treatment/Data/Output/Spectrum_long_grouped_20200914.csv", na="")



```






## look at difference between actual ages - aging in/out actual loss or growth 
#ex actual 2019 / actual 2018

##calc predicted % difference
#predicted 2019/actual 2018

##Compare actual to predicted - expected impact based on aging


```{r load, echo=FALSE}

## BK's solution
## Note: Make sure to install `janitor` and `dplyr` version >= 1.0.0
## This version calculates the perc for all years
df_perc_disagg<- df_long_Grouped_SemiFine_disagg %>%
  janitor::clean_names() %>% 
  arrange( actual_age_cat_semi_fine) %>% #making year values sequential 
  group_by(country, actual_age_cat_semi_fine) %>% #dont want to confuse the countries or hiv/art obs - keep lag/lead in these groups
  mutate( total_prev_value_actual = lag(total_value_actual),
    
    perc_diff = ifelse(
    !is.na(total_value_actual) & !is.na(total_prev_value_actual), #ignore the values that are empty (na)
    total_value_actual / total_prev_value_actual*100, 
    NA
    ),
    
    perc_of_lastyr = ifelse(
      !is.na(total_value_predict) & !is.na(lag(total_value_actual)),
      total_value_predict / total_prev_value_actual * 100,
      NA
#  ),
 #       perc_diff_Compare2= (perc_diff/perc_of_lastyr)*100
 )) %>% 
  ungroup() %>% 
  relocate(total_prev_value_actual, .after = total_value_actual)
  
View(df_perc_disagg)
  


  

```



#filter ART, time and select relevant variables



```{r load, echo=FALSE}
spectrum_clean_all_disagg<-df_perc_disagg%>%
  filter(time %in% c('2019')) %>%
  select(country, time, sex, actual_age_cat_semi_fine, perc_of_lastyr)%>%
  rename(year=time, age_group=actual_age_cat_semi_fine)%>%
  mutate_if(is.factor, as.character)%>%
  mutate(sex=ifelse(sex==" Female", "Female", sex)) %>%
  mutate(sex=ifelse(sex==" Male", "Male", sex)) %>%
  glimpse()

view(spectrum_clean_all_disagg)
glimpse(spectrum_clean_all_disagg)


write_csv(spectrum_clean_all_disagg,"C:/Users/jesse/OneDrive/Documents/Work Material/USAID/Treatment/Data/Output/spectrum_clean_all_total_20201123_FY20.csv", na="")

```







#################################   ARCHIVE CODE ##############################

<!-- ## Start with long non-disaggregated because it has all age groups "MaleFemale" -->

<!-- #### SKIP BELOW FOR SEX DISAGGREGATE DATA OR ALL THE WAY TO BOTTOM TOTAL DATASET (MALEFEMALE, MALE, FEMALE) -->


<!-- ########################################################### -->
<!-- ######    NO DISAGGREGATE (format 1) ###### -->
<!-- ########################################################### -->

<!-- ```{r load, echo=FALSE} -->

<!-- #2 -->
<!--   #drop the malefemale==0 and sex -->

<!-- df_long_total <- df_full_PREDICTagegroups %>% -->
<!--     subset(sex="MaleFemale") %>% -->
<!--     select(!sex) %>% -->
<!--     glimpse() -->


<!-- # df_long_total <- df_full_PREDICTagegroups %>% -->
<!-- #     subset(malefemale==1) %>% -->
<!-- #     select(!malefemale) %>% -->
<!-- #     select(!sex) %>% -->
<!-- #     glimpse() -->
<!-- #    -->
<!-- view(df_long_total) -->
<!-- glimpse(df_long_total) -->

<!-- #Rows: 45,360 -->
<!-- #Columns: 9 -->
<!-- #  -->
<!-- # write_csv(df_long_total,"C:/Users/jesse/OneDrive/Documents/Work Material/USAID/Treatment/Data/Output/Spectrum_long_20200914.csv", na="") -->
<!-- ``` -->



<!-- ## Group by Age Groups and Country  -->
<!-- #actual and predicted separately -->
<!-- #merge actual and predicted -->

<!-- ```{r load, echo=FALSE} -->

<!-- #insert the df_long_total -->

<!-- #actual -->
<!-- df_long_Grouped_Actual_SemiFine<- df_long_total %>% -->
<!--   group_by(Country, Time, ACTUAL_Age_cat_SemiFine) %>% -->
<!--   summarise(Total_Value_Actual=sum(Value)) -->

<!-- glimpse(df_long_Grouped_Actual_SemiFine) -->
<!-- view(df_long_Grouped_Actual_SemiFine) -->

<!-- #predict -->
<!-- df_long_Grouped_Predict_SemiFine<- df_long_total %>% -->
<!--   group_by(Country, Time_predict, PREDICT_Age_cat_SemiFine) %>% -->
<!--   summarise(Total_Value_Predict=sum(Value)) -->

<!-- glimpse(df_long_Grouped_Predict_SemiFine) -->
<!-- view(df_long_Grouped_Predict_SemiFine) -->


<!-- #rename predict variables to match actual so they can merge -->
<!-- df_long_Grouped_Predict_SemiFine_vars<- df_long_Grouped_Predict_SemiFine %>% -->
<!--   rename(Time=Time_predict, ACTUAL_Age_cat_SemiFine=PREDICT_Age_cat_SemiFine) -->

<!-- glimpse(df_long_Grouped_Predict_SemiFine_vars) -->
<!-- view(df_long_Grouped_Predict_SemiFine_vars) -->

<!-- #merge back together -->
<!-- df_long_Grouped_SemiFine <- df_long_Grouped_Actual_SemiFine %>%  -->
<!--   left_join(df_long_Grouped_Predict_SemiFine_vars, by=c("Time"="Time", "ACTUAL_Age_cat_SemiFine"="ACTUAL_Age_cat_SemiFine","Country"="Country")) -->

<!-- view(df_long_Grouped_SemiFine) -->





<!-- ``` -->






<!-- ## look at difference between actual ages - aging in/out actual loss or growth  -->
<!-- #ex actual 2019 / actual 2018 -->

<!-- ##calc predicted % difference -->
<!-- #predicted 2019/actual 2018 -->

<!-- ##Compare actual to predicted - expected impact based on aging -->


<!-- ```{r load, echo=FALSE} -->

<!-- ## BK's solution -->
<!-- ## Note: Make sure to install `janitor` and `dplyr` version >= 1.0.0 -->
<!-- ## This version calculates the perc for all years -->
<!-- df_perc<- df_long_Grouped_SemiFine %>% -->
<!--   janitor::clean_names() %>%  -->
<!--   arrange(actual_age_cat_semi_fine) %>% #making year values sequential  -->
<!--   group_by(country, actual_age_cat_semi_fine) %>% #dont want to confuse the countries or hiv/art obs - keep lag/lead in these groups -->
<!--   mutate( total_prev_value_actual = lag(total_value_actual), -->

<!--     perc_diff = ifelse( -->
<!--     !is.na(total_value_actual) & !is.na(total_prev_value_actual), #ignore the values that are empty (na) -->
<!--     total_value_actual / total_prev_value_actual*100,  -->
<!--     NA -->
<!--     ), -->

<!--     perc_of_lastyr = ifelse( -->
<!--       !is.na(total_value_predict) & !is.na(lag(total_value_actual)), -->
<!--       total_value_predict / total_prev_value_actual * 100, -->
<!--       NA -->
<!-- #  ), -->
<!--  #       perc_diff_Compare2= (perc_diff/perc_of_lastyr)*100 -->
<!--  )) %>%  -->
<!--   ungroup() %>%  -->
<!--   relocate(total_prev_value_actual, .after = total_value_actual) -->

<!-- View(df_perc) -->





<!-- ``` -->



<!-- #filter ART, time and select relevant variables -->


<!-- ```{r load, echo=FALSE} -->
<!-- spectrum_clean_all<-df_perc%>% -->
<!--   filter(time %in% c( '2019')) %>% -->
<!--   select(country, time, actual_age_cat_semi_fine,   perc_of_lastyr)%>% -->
<!--     rename(year=time, age_group=actual_age_cat_semi_fine)%>% -->
<!--     mutate_if(is.factor, as.character)%>% -->
<!--   glimpse() -->

<!-- view(spectrum_clean_all) -->
<!-- write_csv(spectrum_clean_all,"C:/Users/jesse/OneDrive/Documents/Work Material/USAID/Treatment/Data/Output/spectrum_clean_all_20200916.csv", na="") -->

<!-- ``` -->










<!-- ########################################################### -->
<!-- ######    SEX DISAGGREGATE (format 2) ###### -->
<!-- ########################################################### -->


<!-- Multiple data formats -->
<!-- 1) long with disaggregate -->
<!-- 2) long without sex disaggregate -->
<!-- 3) long with all data (male, female and malefemale) - can be disaggregated later -->
<!-- ```{r load, echo=FALSE} -->


<!-- #1 -->
<!--   #drop the malefemale==1 -->
<!-- df_long_sexdisagg <- df_full_PREDICTagegroups %>% -->
<!--     subset(sex!=2) %>% -->
<!--     view() -->

<!-- view(df_long_sexdisagg) -->
<!-- glimpse(df_long_sexdisagg) -->
<!-- #Rows: 73,920 -->
<!-- #Columns: 10 -->

<!-- # write_csv(df_long_sexdisagg,"C:/Users/jesse/OneDrive/Documents/Work Material/USAID/Treatment/Data/Output/Spectrum_long_sexdisagg_20200914.csv", na="") -->

<!-- ``` -->





<!-- ## Group by Age Groups and Country and sex -->
<!-- #actual and predicted separately -->
<!-- #merge actual and predicted -->

<!-- ```{r load, echo=FALSE} -->

<!-- #insert the df_long_sexdisagg -->

<!-- #actual -->
<!-- df_long_Grouped_Actual_SemiFine_disagg<- df_long_sexdisagg %>% -->
<!--   group_by(Country, sex, Time, ACTUAL_Age_cat_SemiFine) %>% -->
<!--   summarise(Total_Value_Actual=sum(Value)) -->

<!-- glimpse(df_long_Grouped_Actual_SemiFine_disagg) -->
<!-- view(df_long_Grouped_Actual_SemiFine_disagg) -->

<!-- #predict -->
<!-- df_long_Grouped_Predict_SemiFine_disagg<- df_long_sexdisagg %>% -->
<!--   group_by(Country, sex, Time_predict, PREDICT_Age_cat_SemiFine) %>% -->
<!--   summarise(Total_Value_Predict=sum(Value)) -->

<!-- glimpse(df_long_Grouped_Predict_SemiFine_disagg) -->
<!-- view(df_long_Grouped_Predict_SemiFine_disagg) -->


<!-- #rename predict variables to match actual so they can merge -->
<!-- df_long_Grouped_Predict_SemiFine_vars_disagg<- df_long_Grouped_Predict_SemiFine_disagg %>% -->
<!--   rename(Time=Time_predict, ACTUAL_Age_cat_SemiFine=PREDICT_Age_cat_SemiFine) -->

<!-- glimpse(df_long_Grouped_Predict_SemiFine_vars_disagg) -->
<!-- view(df_long_Grouped_Predict_SemiFine_vars_disagg) -->

<!-- #merge back together -->
<!-- df_long_Grouped_SemiFine_disagg <- df_long_Grouped_Actual_SemiFine_disagg %>%  -->
<!--   left_join(df_long_Grouped_Predict_SemiFine_vars_disagg, by=c("Time"="Time", "ACTUAL_Age_cat_SemiFine"="ACTUAL_Age_cat_SemiFine","Country"="Country", "sex"="sex")) -->

<!-- view(df_long_Grouped_SemiFine_disagg) -->





<!-- ``` -->






<!-- ## look at difference between actual ages - aging in/out actual loss or growth  -->
<!-- #ex actual 2019 / actual 2018 -->

<!-- ##calc predicted % difference -->
<!-- #predicted 2019/actual 2018 -->

<!-- ##Compare actual to predicted - expected impact based on aging -->


<!-- ```{r load, echo=FALSE} -->

<!-- ## BK's solution -->
<!-- ## Note: Make sure to install `janitor` and `dplyr` version >= 1.0.0 -->
<!-- ## This version calculates the perc for all years -->
<!-- df_perc_disagg<- df_long_Grouped_SemiFine_disagg %>% -->
<!--   janitor::clean_names() %>%  -->
<!--   arrange( actual_age_cat_semi_fine) %>% #making year values sequential  -->
<!--   group_by(country, actual_age_cat_semi_fine) %>% #dont want to confuse the countries or hiv/art obs - keep lag/lead in these groups -->
<!--   mutate( total_prev_value_actual = lag(total_value_actual), -->

<!--     perc_diff = ifelse( -->
<!--     !is.na(total_value_actual) & !is.na(total_prev_value_actual), #ignore the values that are empty (na) -->
<!--     total_value_actual / total_prev_value_actual*100,  -->
<!--     NA -->
<!--     ), -->

<!--     perc_of_lastyr = ifelse( -->
<!--       !is.na(total_value_predict) & !is.na(lag(total_value_actual)), -->
<!--       total_value_predict / total_prev_value_actual * 100, -->
<!--       NA -->
<!-- #  ), -->
<!--  #       perc_diff_Compare2= (perc_diff/perc_of_lastyr)*100 -->
<!--  )) %>%  -->
<!--   ungroup() %>%  -->
<!--   relocate(total_prev_value_actual, .after = total_value_actual) -->

<!-- View(df_perc_disagg) -->





<!-- ``` -->



<!-- #filter ART, time and select relevant variables -->


<!-- ```{r load, echo=FALSE} -->
<!-- spectrum_clean_disagg<-df_perc_disagg%>% -->
<!--   filter(time %in% c('2019')) %>% -->
<!--   select(country, time, sex, actual_age_cat_semi_fine, perc_of_lastyr)%>% -->
<!--   rename(year=time, age_group=actual_age_cat_semi_fine)%>% -->
<!--   mutate_if(is.factor, as.character)%>% -->
<!--   mutate(sex=ifelse(sex==" Female", "Female", "Male")) %>% -->
<!--   glimpse() -->

<!-- view(spectrum_clean_disagg) -->
<!-- glimpse(spectrum_clean_disagg) -->

<!-- write_csv(spectrum_clean_disagg,"C:/Users/jesse/OneDrive/Documents/Work Material/USAID/Treatment/Data/Output/spectrum_clean_sexdisagg_20200916.csv", na = "") -->

<!-- ``` -->













### Aditional Archives


<!-- # Figure out predicted ages  -->
<!-- #ex, 2017 ages if 2016 aged w/o other loss/gains -->


<!-- ```{r load, echo=FALSE} -->



<!-- PREDICTagebreaks <- c(0,4,9,14,19,24,29,34,39,44,49,85) -->
<!-- PREDICTagelabels <- c("0-3","4-8","9-13","14-18","19-23","24-28","29-33", -->
<!--                "34-38","39-43","44-48","49+") -->

<!-- PREDICTagebreaks18 <- c(0,9,14,19,24,29,34,39,44,49,85) -->
<!-- PREDICTagelabels18 <- c("0-8","9-13","14-18","19-23","24-28","29-33", -->
<!--                "34-38","39-43","44-48","49+") -->

<!-- df_wide_PREDICTagegroupsFINE <- setDT(df_combined_PEPFARagegroups) [ , PREDICT_Age_cat_Fine := cut(Age_num,  -->
<!--                                 breaks = PREDICTagebreaks,  -->
<!--                                 right = FALSE,  -->
<!--                                 labels = PREDICTagelabels)] -->
<!-- view(df_wide_PREDICTagegroupsFINE) -->
<!-- glimpse(df_wide_PREDICTagegroupsFINE) -->

<!-- df_wide_PREDICTagegroupsALL <- setDT(df_wide_PREDICTagegroupsFINE) [ , PREDICT_Age_cat_SemiFine := cut(Age_num,  -->
<!--                                 breaks = PREDICTagebreaks18,  -->
<!--                                 right = FALSE,  -->
<!--                                 labels = PREDICTagelabels18)] -->
<!-- view(df_wide_PREDICTagegroupsALL) -->
<!-- glimpse(df_wide_PREDICTagegroupsALL) -->

<!-- ``` -->



<!-- ## Figure out predicted ages  -->
<!-- #ex, 2017 ages if 2016 aged w/o other loss/gains -->


<!-- ```{r load, echo=FALSE} -->


<!-- for each year and actual age cat -->
<!--   do value of previous year and  -->


<!-- PRED_Value_2019(ACTUAL_Age_cat_Fine) = ACTUAL_Value_2018 (PRED_age_bands) -->


<!-- #loop over age bands -->

<!-- #long -->

<!-- for(i in Value){ -->
<!--   Pred_Value <- when Time=Time-1 & PRED_Age_cat_SemiFine=0 -->



<!--  # wide??? -->

<!--   sum(ACTUAL_Value_2018) if PRED_age_bands =  -->
<!-- } -->

<!-- PRED_Value_2019_1-9  <- ACTUAL_Value_2018_0-8 -->

<!-- PRED_Value_2019_1-9  <- ACTUAL_Value_2018_0-8 -->


<!-- ``` -->



